Trauma treatment delay

### 2023-09-21 manuscriptlongversion.Rmd
intro

Within modern hospital care, trauma care stands as an important part in healthcare and is crucial to lowering mortality and morbidity in injured patients<!-- Not only hospital trauma care, you may want to mention the entire system already here-->. The definition of trauma is "a set of psychological and physiological responses to pain, injury, serious illness, medical procedures and frightening treatment experiences" [@International_Society_for_Traumatic_Stress_Studies_2020]. As the definition suggests, it encompasses a broad spectrum of different conditions, both physical and psychological<!-- I don't think you need this definition here, it doesn't really fit-->. The one we will be focusing on in this paper is physical trauma, the clinical entity composed of physical injury and the body's associated response, hereby referred to as trauma.

@@

It is often divided into two subgroups<!-- By "it", do you mean trauma? Suggest you just write so-->, blunt force trauma and penetrating trauma. Blunt force trauma is when an object or force strikes the body, often causing bruising, broken bones or deep cuts. Examples of this could be car crashes, falls or direct blows to the body. Penetrating<!-- More common to refer to it as penetrating, rather than penetrative--> trauma is when an object pierces the skin or body and creates an open wound. Examples of this<!-- By "this", do you mean penetrating trauma mechanisms?--> are gunshot wounds and stab wounds [@National_Institute_of_General_Medical_Sciences_2020]. Differences exist between the two groups, as blunt trauma patients tend to score higher on ISS scale<!-- You haven't mentioned ISS before here. Do you mean that blunt trauma patients are more severely injured?--> on arrival and as such require more resources and be hospitalized for a longer period because of their more severe injuries.

Trauma is and has long been a major cause of death around the world today, taking the lives of around 4,4 million people each year, almost 8% of all deaths [@World_Health_Organization_2021]. In the United States, trauma is the 4th leading cause of death among the general population and the leading cause for people between the ages of 1 and 44 [@AAST_2020]. It is also an important cause of hospitalization and morbidity among all age groups, including seniors, and is responsible for an estimated 10% of all years lived with a disability globally. This has a significant burden on social and economic level, costing countries billions of US dollars each year in healthcare and lost productivity[@connolly_woo_lampron_perry_2018]. Studies estimate the cost of trauma care being between 18,500\$ and 41,500\$ in HIC<!-- Spell out the first time you mention HIC-->, depending on the country [@Willenberg_Curtis_Taylor_Jan_Glass_Myburgh_2012]. However, as one might expect<!-- Phrases like "as one might expect", "it is known that", "as has been shown" are usually unecessary and don't add much. Only use them if you need the extra words -->, this burden is not evenly distributed between or within countries. Many social factors such as age, sex and social status play a major role in the risk of dying from trauma, with young men with low socio-economic status being at most risk [@World_Health_Organization_2021]. It is not only patient level factors that affect how people might be affected<!-- affect and affect--> by trauma. About 90% of all trauma related deaths occur in low- and middle-income countries (LMIC), with death rates by trauma also being higher than in high income countries. Even within these LMIC, people of poorer socio-economic status have higher death rates from trauma. Problems identified within these countries that contribute to these statisitics were infrastructure, education and training, attributed to lack of funding, brain drain to HIC and lack of availability of basic amenities [@Shanthakumar_Payne_Leitch_Alfa-Wali_2021]. In HIC where funding and governance over the healthcare system is better designed, functioning trauma systems and dedicated trauma centers exist that have been shown to lower mortality but also improve functional outcome in trauma patients [@Nirula_Brasel_2006].

>>>>

Trauma, defined as the clinical entity composed of physical injury and the body's associated response, is and has long been a major cause of death around the world today.

Within modern hospital care, trauma systems stands as an important part in healthcare and is crucial to lowering mortality and morbidity in injured patients through many means, such as hospital care, follow up and prevention programs.

@@

Trauma is often divided into two subgroups, blunt force trauma and penetrating trauma. Blunt force trauma is when an object or force strikes the body, often causing bruising, broken bones or deep cuts. Examples of blunt trauma could be car crashes, falls or direct blows to the body. Penetrating trauma is when an object pierces the skin or body and creates an open wound. Examples of penetrating trauma are gunshot wounds and stab wounds [@National_Institute_of_General_Medical_Sciences_2020]. Differences exist between the two groups, as blunt trauma patients tend to be more injured on arrival to the hospital and as such require more resources and be hospitalized for a longer period because of their more severe injuries [@FITCH20191201].

Today, trauma takes the lives of around 4,4 million people each year, almost 8% of all deaths [@World_Health_Organization_2021]. In the United States, trauma is the 4th leading cause of death among the general population and the leading cause for people between the ages of 1 and 44 [@AAST_2020]. It is also an important cause of hospitalization and morbidity among all age groups, including seniors, and is responsible for an estimated 10% of all years lived with a disability globally. This has a significant burden on social and economic level, costing countries billions of US dollars each year in healthcare and lost productivity[@connolly_woo_lampron_perry_2018]. Studies estimate the cost of trauma care being between 18,500\$ and 41,500\$ in high income countries, depending on the country [@Willenberg_Curtis_Taylor_Jan_Glass_Myburgh_2012]. However, as one might expect<!-- Phrases like "as one might expect", "it is known that", "as has been shown" are usually unecessary and don't add much. Only use them if you need the extra words -->, this burden is not evenly distributed between or within countries. Many social factors such as age, sex and social status play a major role in the risk of dying from trauma, with young men with low socio-economic status being at most risk [@World_Health_Organization_2021]. It is not only patient level factors that affect how people might be affected<!-- affect and affect--> by trauma. About 90% of all trauma related deaths occur in low- and middle-income countries (LMIC), with death rates by trauma also being higher than in high income countries. Even within these LMIC, people of poorer socio-economic status have higher death rates from trauma. Problems identified within these countries that contribute to these statisitics were infrastructure, education and training, attributed to lack of funding, brain drain to HIC and lack of availability of basic amenities [@Shanthakumar_Payne_Leitch_Alfa-Wali_2021]. In HIC where funding and governance over the healthcare system is better designed, functioning trauma systems and dedicated trauma centers exist that have been shown to lower mortality but also improve functional outcome in trauma patients [@Nirula_Brasel_2006].
